### 2024-09-17 base-rates.qmd
20240917 - nomogram for Nate Silver examples

1. the [conditional probability](#sec-conditionalProbability) the likelihood of observing the evidence, $R$, given that the hypothesis, $C$, is true (i.e., $P(R|C)$; [sensitivity](#sec-sensitivity))
1. the [conditional probability](#sec-conditionalProbability) the likelihood of observing the evidence, $R$, given that the hypothesis, $C$, is false (i.e., $P(R | \text{not } C)$; [false positive rate](#sec-falsePositiveRate))
1. the [marginal probability](#sec-baseRate) (base rate) of the event occurring (i.e., the prior probability of the hypothesis, $C$, being true; $P(C_i)$)

@@

@@

@@

@@

The function can also estimate the posttest probability of an event given the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN):

@@

We discuss true positives (TP), true negatives (TN), false positives (FP), false negatives (FN), sensitivity (SN), and specificity (SP) in INSERT.

@@

A probability nomogram calculator can be found at the following link: [http://araw.mede.uic.edu/cgi-bin/testcalc.pl](http://araw.mede.uic.edu/cgi-bin/testcalc.pl) (archived at <https://perma.cc/X8TF-7YBX>).
The [`petersenlab`](https://github.com/DevPsyLab/petersenlab) package [@R-petersenlab] contains the `nomogrammer()` function that creates a nomogram plot using the [positive](#sec-positiveLikelihoodRatio) and [negative](#sec-negativeLikelihoodRatio) [likelihood ratio](#sec-diagnosticLikelihoodRatio) or using the sensitivity (SN) and specificity (SP) of the test, as adapted from Adam Chekroud (<https://github.com/achekroud/nomogrammer>):

@@

The function can also create a nomogram plot using the true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN):

@@

The function can also create a nomogram plot using the sensitivity (SN) and selection rate of the test.
Here is a nomogram plot from the [HIV example](#sec-baseRatesOverview):

@@

Here is a nomogram plot from the [cab example](#sec-cabExample):

@@

>>>>

1. the [conditional probability](#sec-conditionalProbability) the likelihood of observing the evidence, $R$, given that the hypothesis, $C$, is true (i.e., $P(R|C)$; [true positive rate](#sec-sensitivity))
1. the [conditional probability](#sec-conditionalProbability) the likelihood of observing the evidence, $R$, given that the hypothesis, $C$, is false (i.e., $P(R | \text{not } C)$; [false positive rate](#sec-falsePositiveRate))
1. the [marginal probability](#sec-baseRate) (base rate) of the event occurring (i.e., the prior probability of the hypothesis, $C$, being true; $P(C_i)$)

Thus, the formula uses the [base rate](#sec-baseRate), the [true positive rate](#sec-sensitivity) (sensitivity), and the [false positive rate](#sec-falsePositiveRate).
The ratio of the [true positive rate](#sec-sensitivity) to the [false positive rate](#sec-falsePositiveRate) is called the [positive likelihood ratio](#sec-positiveLikelihoodRatio), and is used in [Bayesian updating](#sec-bayesianUpdating).

### Example 1: Is Your Partner Cheating on You? {#sec-nateSilverExample1}

@@

### Example 2: Does a Woman Have Cancer? {#sec-nateSilverExample2}

@@

### Example 3: Was it a Terrorist Attack? {#sec-nateSilverExample3}

#### Example 3A {#sec-nateSilverExample3A}

@@

#### Example 3B {#sec-nateSilverExample3B}

@@

The function can also estimate the posttest probability of an event given the number of [true positives](#sec-truePositive) (TP), [true negatives](#sec-trueNegative) (TN), [false positives](#sec-falsePositive) (FP), and [false negatives](#sec-falseNegative) (FN):

@@

We discuss [true positives](#sec-truePositive) (TP), [true negatives](#sec-trueNegative) (TN), [false positives](#sec-falsePositive) (FP), [false negatives](#sec-falseNegative) (FN), [sensitivity](#sec-sensitivity) (SN), and [specificity](#sec-specificity) (SP) in @sec-thresholdDependentAccuracy (@sec-decisionOutcomes and @sec-sensitivitySpecificityPPVnpv).

@@

A probability nomogram calculator can be found at the following link: [http://araw.mede.uic.edu/cgi-bin/testcalc.pl](http://araw.mede.uic.edu/cgi-bin/testcalc.pl) (archived at <https://perma.cc/X8TF-7YBX>).
The [`petersenlab`](https://github.com/DevPsyLab/petersenlab) package [@R-petersenlab] contains the `nomogrammer()` function that creates a nomogram plot using the [positive](#sec-positiveLikelihoodRatio) and [negative](#sec-negativeLikelihoodRatio) [likelihood ratio](#sec-diagnosticLikelihoodRatio) or using the [sensitivity](#sec-sensitivity) (SN) and [specificity](#sec-specificity) (SP) of the test, as adapted from Adam Chekroud (<https://github.com/achekroud/nomogrammer>):

@@

The function can also create a nomogram plot using the [true positives](#sec-truePositive) (TP), [true negatives](#sec-trueNegative) (TN), [false positives](#sec-falsePositive) (FP), and [false negatives](#sec-falseNegative) (FN):

@@

The function can also create a nomogram plot using the [sensitivity](#sec-sensitivity) (SN) and [selection rate](#sec-selectionRatio) of the test.
Here is a nomogram plot from the [HIV example](#sec-baseRatesOverview):

@@

Here is a nomogram plot from the [cab example](#sec-cabExample) [@Kahneman2011]:

@@

Here is a nomogram plot from [Example 1](#sec-nateSilverExample1) from @Silver2012 (where [specificity](#sec-specificity) equals 1 minus the [false positive rate](#sec-falsePositiveRate)):

```{r}
petersenlab::nomogrammer(
  pretestProb = .04,
  SN = .50,
  SP = 1 - .05
)
```

Here is a nomogram plot from [Example 2](#sec-nateSilverExample2) from @Silver2012 (where [specificity](#sec-specificity) equals 1 minus the [false positive rate](#sec-falsePositiveRate)):

```{r}
petersenlab::nomogrammer(
  pretestProb = .014,
  SN = .75,
  SP = 1 - .10
)
```

Here is a nomogram plot from [Example 3A](#sec-nateSilverExample3A) from @Silver2012 (where [specificity](#sec-specificity) equals 1 minus the [false positive rate](#sec-falsePositiveRate)):

```{r}
petersenlab::nomogrammer(
  pretestProb = .00005,
  SN = 1,
  SP = 1 - .00008
)
```

Here is a nomogram plot from [Example 3B](#sec-nateSilverExample3B) from @Silver2012 (where [specificity](#sec-specificity) equals 1 minus the [false positive rate](#sec-falsePositiveRate)):

```{r}
petersenlab::nomogrammer(
  pretestProb = .3846272,
  SN = 1,
  SP = 1 - .00008
)
```
